---
title: "Paper Example 1: Linked-Test Design with 1PL Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paper Example 1: Linked-Test Design with 1PL Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Purpose

This vignette is a **faithful reproduction** of Example 1 from
Schroeders and Gnambs (2025), *"Sample Size Planning for Item Response
Models: A Tutorial for the Quantitative Researcher"* and its companion
R code at
<https://ulrich-schroeders.github.io/IRT-sample-size/example_1.html>.
The goal is to let irtsim users compare its Monte Carlo output directly
against the published reference.

The paper's Example 1 asks a classic question: **how many examinees are
needed to recover item difficulty parameters in a linked, two-form
achievement test fit with a Rasch (1PL) model?**

```{r load, include = FALSE}
library(irtsim)
library(ggplot2)

ex1_path <- if (file.exists("../inst/extdata/vignette_ex1_paper.rds")) {
  "../inst/extdata/vignette_ex1_paper.rds"
} else {
  system.file("extdata", "vignette_ex1_paper.rds", package = "irtsim")
}

if (!nzchar(ex1_path)) {
  stop(
    "Precomputed results not found. ",
    "Run data-raw/precompute_vignettes.R before building this vignette."
  )
}

ex1 <- readRDS(ex1_path)
res <- ex1$results
```

## Design, from the paper

| Decision | Paper value | irtsim mapping |
|---|---|---|
| Estimation model | 1PL (Rasch) | `estimation_model = "1PL"` |
| Number of items | 30 | `n_items = 30` |
| Item discriminations (generation) | `rnorm(30, 1, 0.1)` | `item_params$a` |
| Item difficulties | `seq(-2, 2, length.out = 30)` | `item_params$b` |
| Two forms, common block items 13–18 | 2 × 30 linking matrix | `missing = "linking"` |
| Monte Carlo iterations | 438 | `iterations = 438` |
| Sample sizes | `seq(100, 600, 50)` | `sample_sizes` |
| Performance criterion | MSE, threshold 0.05 | `summary(res)$item_summary$mse` |

Note on the data-generating model: the paper uses a near-constant
discrimination (mean 1, sd 0.1) and fits a Rasch model. irtsim's 1PL
generation fixes `a = 1` exactly, so to match the paper we generate
under a 2PL with `a ~ rnorm(30, 1, 0.1)` and set `estimation_model =
"1PL"`. The estimation model is the one the paper targets; the generation
model is a faithful implementation of the paper's `a` draws.

## Reproducing the study

The code below mirrors the paper. It is shown for reference; the actual
simulation is precomputed and cached in
`inst/extdata/vignette_ex1_paper.rds` to keep vignette build time low.

```{r code, eval = FALSE}
library(irtsim)

set.seed(2024)
n_items <- 30L

# Item parameters exactly as in the paper
a_vals <- rnorm(n_items, mean = 1, sd = 0.1)
b_vals <- seq(-2, 2, length.out = n_items)

# Linking matrix: form 1 = odd items + common block 13-18,
#                 form 2 = even items + common block 13-18
linking_matrix <- matrix(0L, nrow = 2L, ncol = n_items)
linking_matrix[1L, sort(unique(c(seq(1L, n_items, 2L), 13:18)))] <- 1L
linking_matrix[2L, sort(unique(c(seq(2L, n_items, 2L), 13:18)))] <- 1L

design <- irt_design(
  model       = "2PL",
  n_items     = n_items,
  item_params = list(a = a_vals, b = b_vals),
  theta_dist  = "normal"
)

study <- irt_study(
  design,
  sample_sizes     = seq(100L, 600L, by = 50L),
  missing          = "linking",
  test_design      = list(linking_matrix = linking_matrix),
  estimation_model = "1PL"
)

res <- irt_simulate(
  study,
  iterations = 438L,
  seed       = 2024L,
  parallel   = TRUE
)
```

## Summary: MSE by sample size

We summarize the recovered item-difficulty MSE and pair it with its
Monte Carlo standard error (MCSE), following Morris et al. (2019).

```{r summary}
s <- summary(res, criterion = c("mse", "mcse_mse"), param = "b")
head(s$item_summary, 10)
```

The paper plots the MSE trajectory for two representative items: **item
1** (difficulty ≈ −2, extreme) and **item 15** (difficulty ≈ 0,
central). We do the same.

```{r plot}
item_df <- s$item_summary
focal <- subset(item_df, item %in% c(1L, 15L))
focal$item_label <- factor(
  focal$item,
  levels = c(1L, 15L),
  labels = c("Item 1 (b \u2248 -2)", "Item 15 (b \u2248 0)")
)

ggplot(focal, aes(x = sample_size, y = mse, colour = item_label)) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = pmax(mse - 1.96 * mcse_mse, 0),
      ymax = mse + 1.96 * mcse_mse
    ),
    width = 15
  ) +
  scale_x_continuous(breaks = seq(100, 600, 100)) +
  labs(
    title    = "Example 1: MSE of b-parameter vs. sample size",
    subtitle = "Dashed line = paper's 0.05 MSE threshold",
    x        = "Sample size (N)",
    y        = "MSE(b)",
    colour   = NULL
  ) +
  theme_minimal(base_size = 12)
```

## Recommended N

Using irtsim's built-in `recommended_n()` helper we can extract the
smallest N that meets the paper's MSE ≤ 0.05 threshold for each item.
The paper reports sample-size requirements in the same sense.

```{r rec_n}
sim_summary <- summary(res, criterion = "mse", param = "b")
rec <- recommended_n(sim_summary, criterion = "mse", threshold = 0.05,
                     param = "b", aggregate = "none")
head(rec, 10)
```

## Comparison notes — paper vs. irtsim

**Should reproduce (within MC noise):**

- MSE trajectory shape for central items (e.g., item 15) is tightest.
- Extreme items (item 1, item 30) may not reach the 0.05 threshold
  within the tested sample-size range (N ≤ 600). An `NA` in the
  `recommended_n` column means the criterion was never met — this is
  informative, not an error.
- Required N to meet MSE ≤ 0.05 for the central block (items 13–18) is
  substantially smaller than for the tails.

**Expected small numerical differences:**

1. **Form assignment.** The paper uses random per-examinee form
   assignment (`sample(c(1, 2), n, replace = TRUE)`). irtsim's
   `apply_missing_structured()` uses deterministic round-robin
   assignment. At N ≥ 100 the induced difference in per-form sample
   sizes is small (< 1 examinee gap) and does not materially shift the
   MSE trajectories.
2. **RNG dispatch.** irtsim runs in parallel mode with
   `future.seed = TRUE`, which uses L'Ecuyer-CMRG substreams. The paper
   uses the session default Mersenne-Twister. Both are valid Monte
   Carlo streams; the specific numbers differ. Only trajectory shape is
   expected to match, not bit-for-bit numerical equality.
3. **Iterations.** We use the paper's 438 iterations as-is. That value
   reflects a Burton (2003) precision target that irtsim exposes via
   [`irt_iterations()`]; users can recompute it if they want a
   different MC precision.

## References

Burton, A., Altman, D. G., Royston, P., & Holder, R. L. (2006). The
design of simulation studies in medical statistics. *Statistics in
Medicine*, 25, 4279–4292. <https://doi.org/10.1002/sim.2673>

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation
studies to evaluate statistical methods. *Statistics in Medicine*, 38,
2074–2102. <https://doi.org/10.1002/sim.8086>

Schroeders, U., & Gnambs, T. (2025). Sample size planning for item
response models: A tutorial for the quantitative researcher.
Companion code: <https://ulrich-schroeders.github.io/IRT-sample-size/>.
